Exact asymptotic expressions for local clustering in hyperbolic random graphs

### 2018-10-17 clustering_infinite_model.tex
small update on motivation and figure 3

>>>>

\begin{figure}[!t]
\centering
\begin{tikzpicture}
	%Define the coordinates 
	%p = (\u,\v) and p^\prime = (\uu, \vv)
	%Box \Rcal_n has width 2\r and height \t
	\pgfmathsetmacro{\u}{0} %0
	\pgfmathsetmacro{\v}{0.5} %1
	\pgfmathsetmacro{\vv}{1.2} %0.8
	\pgfmathsetmacro{\vvv}{2} %0.8
	\pgfmathsetmacro{\uuu}{-3} %1.4	
	\pgfmathsetmacro{\r}{6.5}
	\pgfmathsetmacro{\t}{4}

	\pgfmathsetmacro{\ubound}{exp((\vv+\vvv)/2)-exp((\v+\vvv)/2)}

	\pgfmathsetmacro{\uu}{3*\ubound/4}

	\pgfmathsetmacro{\leftintvandvvv}{-exp((\v + \vvv)/2)}
	\pgfmathsetmacro{\rightintvandvvv}{exp((\v + \vvv)/2)}
	\pgfmathsetmacro{\leftintvvandvvv}{\uu-exp((\vv + \vvv)/2)}
	\pgfmathsetmacro{\rightintvvandvvv}{\uu+exp((\vv + \vvv)/2)}

	%The box \Rcal_n
	\draw[line width=1pt,dashed] (-\r,0) -- (\r,0) -- (\r,\t) -- (-\r,\t) -- (-\r,0);

	%Dram all three nodes
    \draw node[fill, circle, inner sep=0pt, minimum size=5pt] (p1) at (\u,\v) {};
    \path (p1)+(-0.2,0.2) node {$p_0$};
    \draw node[fill,blue, circle, inner sep=0pt, minimum size=5pt] (p2) at (\uu,\vv) {};
    \path (p2)+(0.2,0.2) node {\color{blue}$p_1$};

	%Boundaries p_0 = (\u,\v)

	%Right boundary
	\pgfmathsetmacro{\rightbounduv}{\u+exp((\v)/2)}
	\draw[domain=\rightbounduv:\r,smooth,variable=\x,black,line width=1pt] plot (\x, {2*ln(\x)-\v});
    %Left boundary
    \pgfmathsetmacro{\leftbounduv}{\u-exp((\v)/2)}
    \draw[domain=\leftbounduv:-\r,smooth,variable=\x,black,line width=1pt] plot (\x, {2*ln(-\x)-\v});

    %Boundaries p_1 = (\uu,\vv)

    %Right boundary
    \pgfmathsetmacro{\rightbounduuvv}{\uu+exp((\vv)/2)}
    \draw[domain=\rightbounduuvv:\r,smooth,variable=\x,blue,line width=1pt] plot (\x, {2*ln(\x-\uu)-\vv});
%    %Shifted right boundary
%    \pgfmathsetmacro{\shiftrightbounduuvv}{\uu+exp((\vv + \t)/2)-2*\r}
%    \draw[domain=\shiftrightbounduuvv:-\r,smooth,variable=\x,blue,line width=1pt] plot (\x, {2*ln(\x+(2*\r-\uu))-\vv});
    %Left boundary 
    \pgfmathsetmacro{\leftbounduuvv}{\uu-exp((\vv)/2)}
    \draw[domain=\leftbounduuvv:-\r,smooth,variable=\x,blue,line width=1pt] plot (\x, {2*ln(\uu-\x)-\vv});
%    %Shifted left boundary
%    \pgfmathsetmacro{\shiftleftbounduuvv}{\uu-exp((\vv + \t)/2)+2*\r}
%    \draw[domain=\shiftleftbounduuvv:\r,smooth,variable=\x,blue,line width=1pt] plot (\x, {2*ln(2*\r + \uu-\x)-\vv});

	\draw [red,dotted,line width=1pt] (\leftintvandvvv,\vvv) -- (\rightintvandvvv,\vvv);

	\draw [red,dotted,line width=1pt] (\leftintvvandvvv,\vvv) -- (\rightintvvandvvv,\vvv);

	\draw [red,line width=1pt] (\leftintvandvvv,\vvv) -- (\rightintvandvvv,\vvv);

    \draw node[fill,black!60!green, circle, inner sep=0pt, minimum size=5pt] (p2) at (\uuu,\vvv) {};
    \path (p2)+(0.25,-0.2) node {\color{black!60!green}$p_2$};

	%Boundaries p_2 = (\uuu,\vvv)

	%Right boundary
	\pgfmathsetmacro{\rightbounduuuvvv}{\uuu+exp((\vvv)/2)}
	\draw[domain=\rightbounduuuvvv:\r,smooth,variable=\x,black!60!green,line width=1pt] plot (\x, {2*ln(\x-\uuu)-\vvv});
    %Left boundary
    \pgfmathsetmacro{\leftbounduuuvvv}{\uuu-exp((\vvv)/2)}
    \draw[domain=\leftbounduuuvvv:-\r,smooth,variable=\x,black!60!green,line width=1pt] plot (\x, {2*ln(\uuu-\x)-\vvv});

	\draw node[fill,red, circle, inner sep=0pt, minimum size=3pt] at (\leftintvandvvv,\vvv) {};
	\path (\leftintvandvvv,\vvv)+(-0.5,-0.5) node[red] {$-e^{\frac{y_0 + y_2}{2}}$};
	\draw node[fill,red, circle, inner sep=0pt, minimum size=3pt] at (\rightintvandvvv,\vvv) {};
	\path (\rightintvandvvv,\vvv)+(0,-0.5) node[red] {$e^{\frac{y_0 + y_2}{2}}$};

	\draw node[fill,red, circle, inner sep=0pt, minimum size=3pt] at (\leftintvvandvvv,\vvv) {};
	\path (\leftintvvandvvv,\vvv)+(0.75,0.5) node[red] {$x_1 - e^{\frac{y_0 + y_2}{2}}$};
	\draw node[fill,red, circle, inner sep=0pt, minimum size=3pt] at (\rightintvvandvvv,\vvv) {};
	\path (\rightintvvandvvv,\vvv)+(-0.5,0.5) node[red] {$x_1 + e^{\frac{y_0 + y_2}{2}}$};

\end{tikzpicture}\\
\vspace{0.5cm}
\begin{tikzpicture}
	%Define the coordinates 
	%p = (\u,\v) and p^\prime = (\uu, \vv)
	%Box \Rcal_n has width 2\r and height \t
	\pgfmathsetmacro{\u}{0} %0
	\pgfmathsetmacro{\v}{0.5} %1
	\pgfmathsetmacro{\vv}{1.2} %0.8
	\pgfmathsetmacro{\vvv}{1.8} %0.8
	\pgfmathsetmacro{\uuu}{2.5} %1.4	
	\pgfmathsetmacro{\r}{6.5}
	\pgfmathsetmacro{\t}{4}

	\pgfmathsetmacro{\ubound}{exp((\vv+\vvv)/2)-exp((\v+\vvv)/2)}

	\pgfmathsetmacro{\uu}{5*\ubound/4}

	\pgfmathsetmacro{\leftintvandvvv}{-exp((\v + \vvv)/2)}
	\pgfmathsetmacro{\rightintvandvvv}{exp((\v + \vvv)/2)}
	\pgfmathsetmacro{\leftintvvandvvv}{\uu-exp((\vv + \vvv)/2)}
	\pgfmathsetmacro{\rightintvvandvvv}{\uu+exp((\vv + \vvv)/2)}

	%The box \Rcal_n
	\draw[line width=1pt,dashed] (-\r,0) -- (\r,0) -- (\r,\t) -- (-\r,\t) -- (-\r,0);

	%Dram all three nodes
    \draw node[fill, circle, inner sep=0pt, minimum size=5pt] (p1) at (\u,\v) {};
    \path (p1)+(-0.2,0.2) node {$p_0$};
    \draw node[fill,blue, circle, inner sep=0pt, minimum size=5pt] (p2) at (\uu,\vv) {};
    \path (p2)+(0.2,0.2) node {\color{blue}$p_1$};

	%Boundaries p_0 = (\u,\v)

	%Right boundary
	\pgfmathsetmacro{\rightbounduv}{\u+exp((\v)/2)}
	\draw[domain=\rightbounduv:\r,smooth,variable=\x,black,line width=1pt] plot (\x, {2*ln(\x)-\v});
    %Left boundary
    \pgfmathsetmacro{\leftbounduv}{\u-exp((\v)/2)}
    \draw[domain=\leftbounduv:-\r,smooth,variable=\x,black,line width=1pt] plot (\x, {2*ln(-\x)-\v});

    %Boundaries p_1 = (\uu,\vv)

    %Right boundary
    \pgfmathsetmacro{\rightbounduuvv}{\uu+exp((\vv)/2)}
    \draw[domain=\rightbounduuvv:\r,smooth,variable=\x,blue,line width=1pt] plot (\x, {2*ln(\x-\uu)-\vv});
    %Left boundary 
    \pgfmathsetmacro{\leftbounduuvv}{\uu-exp((\vv)/2)}
    \draw[domain=\leftbounduuvv:-\r,smooth,variable=\x,blue,line width=1pt] plot (\x, {2*ln(\uu-\x)-\vv});

	\draw [red,dotted,line width=1pt] (\leftintvandvvv,\vvv) -- (\rightintvandvvv,\vvv);

	\draw [red,dotted,line width=1pt] (\leftintvvandvvv,\vvv) -- (\rightintvvandvvv,\vvv);

	\draw [red,line width=1pt] (\leftintvvandvvv,\vvv) -- (\rightintvandvvv,\vvv);

    \draw node[fill,black!60!green, circle, inner sep=0pt, minimum size=5pt] (p2) at (\uuu,\vvv) {};
    \path (p2)+(-0.2,0.2) node {\color{black!60!green}$p_2$};

	%Boundaries p_2 = (\uuu,\vvv)

	%Right boundary
	\pgfmathsetmacro{\rightbounduuuvvv}{\uuu+exp((\vvv)/2)}
	\draw[domain=\rightbounduuuvvv:\r,smooth,variable=\x,black!60!green,line width=1pt] plot (\x, {2*ln(\x-\uuu)-\vvv});
    %Left boundary
    \pgfmathsetmacro{\leftbounduuuvvv}{\uuu-exp((\vvv)/2)}
    \draw[domain=\leftbounduuuvvv:-\r,smooth,variable=\x,black!60!green,line width=1pt] plot (\x, {2*ln(\uuu-\x)-\vvv});

	\draw node[fill,red, circle, inner sep=0pt, minimum size=3pt] at (\leftintvandvvv,\vvv) {};
	\path (\leftintvandvvv,\vvv)+(-0.5,-0.55) node[red] {$-e^{\frac{y_0 + y_2}{2}}$};
	\draw node[fill,red, circle, inner sep=0pt, minimum size=3pt] at (\rightintvandvvv,\vvv) {};
	\path (\rightintvandvvv,\vvv)+(0,-0.55) node[red] {$e^{\frac{y_0 + y_2}{2}}$};

	\draw node[fill,red, circle, inner sep=0pt, minimum size=3pt] at (\leftintvvandvvv,\vvv) {};
	\path (\leftintvvandvvv,\vvv)+(0.75,0.5) node[red] {$x_1 - e^{\frac{y_0 + y_2}{2}}$};
	\draw node[fill,red, circle, inner sep=0pt, minimum size=3pt] at (\rightintvvandvvv,\vvv) {};
	\path (\rightintvvandvvv,\vvv)+(-0.5,0.5) node[red] {$x_1 + e^{\frac{y_0 + y_2}{2}}$};

\end{tikzpicture}
\caption{Example configuration of two points $p$ and $p^\prime$ for which $\BallPon{p} \cap \BallPon{p^\prime}$ is not a subset of $\BallPo{p} \cap \BallPo{p^\prime}$. The red region indicates the area of points belonging to $\BallPon{p} \cap \BallPon{p^\prime}$ but not to $\BallPo{p} \cap \BallPo{p^\prime}$.}
\end{figure}

### 2018-10-17 introduction.tex
small update on motivation and figure 3

\MS{Some ideas:
\begin{enumerate}
\item Hyperbolic random graphs as an alternative to Euclidean/standard random geometric graphs
\end{enumerate}
}

\PvdH{Add intro on common features of complex networks and modeling of such networks.}
\MS{I would not write a length introduction to complex networks here (because the paper is quite long already and we do not study networks in general in this paper). So, I would remove the first paragraph/list (or incorporate it as a single sentence in the ensuing paragraph). }
The main topological features shared among many complex networks are:
\begin{enumerate}[\upshape 1)]
\item Broad degree distribution (power-law/scale-free)
\item Strong clustering (community structure)
\item Small path lengths (small world phenomenon)
\end{enumerate}

Hyperbolic random graphs were suggested as a suitable model for complex networks by Krioukov et al.~\cite{krioukov2010hyperbolic}. Indeed, in their seminal paper they show that these graphs have a power-law degree distribution and exhibit strong clustering. In addition it was found that hyperbolic random graphs have doubly logarithmic shortest path lengths~\cite{bringmann2016average} and allow for efficient routing algorithms~\cite{bringmann2017greedy}. Many other results regarding hyperbolic random graphs have since been established. These include size of the largest component~\cite{BodeFountoulakisMuller2015},~\cite{fountoulakis2018law}, [More?] and number of cliques and largest clique~\cite{blasius2018cliques}.

The aim of this paper is to study local clustering in hyperbolic random graphs. The first rigorous mathematical analysis of the local clustering coefficient was done by Gugelmann et al.~\cite{gugelmann2012random}. They show~\cite[Theorem 2.1]{gugelmann2012random} that the local clustering coefficient is concentrated around its expectation which is shown to be $\bigT{1}$ as the number of vertices tends to infinity. In particular this results implies that the local clustering coefficient is asymptotically bounded away from zero. The convergence of this coefficient was not proven. In contrast, the global clustering coefficient was shown~\cite{candellero2016clustering} to converge in probability to some constant which can be explicitly stated as an integral expression, see Theorem 1.2 in \cite{candellero2016clustering}. The authors mention, however, that their analysis needs significant modification to be able to deal with convergence of the local clustering coefficient.

@@

@@

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{figures/HyperRGG_N5000L5.eps}
\caption{Example of a Hyperbolic Random Graph $G_{\H,n}(\alpha, \nu)$ with $\alpha = ??$, $\nu = ??$ and $n = 500$.}
\label{fig:H_graph_example}
\end{figure}

@@

Clustering measures for networks consider the fraction of triangles (triples of connected vertices) in the network. 
\MS{Do we need the definition of the global clustering coefficient? Also, I thought it was defined as 3 times number of triangles divided by number of length 2 paths and I do not see yet whether this agrees with the definition below.}
For instance, Let $G$ be a simple graph, with vertex set $V(G)$ and edge set $E(G)$ and denote by $T_G$ the total number of triangles in $G$. Then the global clustering coefficient is defined as
\[
	c_G = \frac{\T_G}{\binom{E(G)}{2}}.
\]
This coefficient measures the fraction of triangles in the network compared to the total possible number of triangles. A different global measure for clustering, \emph{local clustering coefficient} computes the average triangle density per node. To define this function, let $D_G(v)$ denote the degree of node $v$ and $N_G(k)$ the number of nodes with degree $k$ in the graph $G$. In addition let $\T_G(v,v_1,v_2) = \ind{(v,v_1) \in E} \ind{(v, v_2) \in E} \ind{(v_1, v_2) \in E}$ be the indicator that $v, v_1$ and $v_2$ form a triangle and write
\begin{equation}
	\T_G(v) = \sum_{v_1, v_2 \in V(G)} \T_G(v,v_1,v_2),
\end{equation}
to denote the number of triangles in which node $v$ participates. Then the local clustering coefficient is given by
\[
	c_G = \frac{1}{|V(G)|} \sum_{v \in V(G)} \frac{T_G(v)}{\binom{D_G(v)}{2}}.
\]

However, since the local clustering coefficient assigns just one value to the whole network, representing its triangular structure, it is unable to characterize local structural properties involving triangles. The local clustering function, on the other hand, measures the fraction of triangles to which nodes of a given degree belong, compared to the maximum number of triangles in which they could participate. It therefore characterizes the triangular structure of nodes in the networks, based on their degree, and gives a more detailed look at the overall structure of the network.

The local clustering function is formally defined as
\begin{equation}\label{eq:def_local_clustering_general}
	c_{G}(k) = \begin{cases}
		\frac{1}{N_G(k)} \sum_{v \in V(G)}  \ind{D_G(v) = k} \frac{\T_G(v)}{\binom{k}{2}} &\mbox{if } N_G(k) \ge 1\\
		0 &\mbox{else.}
	\end{cases}
\end{equation}

We note that other versions of the local clustering function have been considered, for instance where nodes with degree larger than $k$ are considered instead of nodes with degree exactly $k$. \PvdH{Motivate our choice for this local clustering function.}

>>>>

\MS{I would not write a length introduction to complex networks here (because the paper is quite long already and we do not study networks in general in this paper). So, I would remove the first paragraph/list (or incorporate it as a single sentence in the ensuing paragraph). }
\PvdH{I agree.}

Hyperbolic random graphs were suggested as a suitable model for complex networks by Krioukov et al.~\cite{krioukov2010hyperbolic}, exhibiting the three main characteristics commonly found in complex networks: 
\begin{enumerate}[\upshape 1)]
\item Broad degree distribution (power-law/scale-free),
\item Strong clustering (community structure),
\item Small path lengths (small world phenomenon).
\end{enumerate}
Indeed, in their seminal paper they show that these graphs have a power-law degree distribution and exhibit strong clustering, while Bringmann et al.~\cite{bringmann2016average} showed that hyperbolic random graphs have doubly logarithmic shortest path lengths. Many other results regarding hyperbolic random graphs have since been established. One important feature of these graphs is that they allow for efficient routing algorithms~\cite{bringmann2017greedy}. Further results include the size of the largest component~\cite{BodeFountoulakisMuller2015},~\cite{fountoulakis2018law}, [More?] and number of cliques and largest clique~\cite{blasius2018cliques}.

The aim of this paper is to study local clustering in hyperbolic random graphs. The first rigorous mathematical analysis of the local clustering coefficient was done by Gugelmann et al.~\cite{gugelmann2012random}. They show~\cite[Theorem 2.1]{gugelmann2012random} that the local clustering coefficient\footnote{Note that in~\cite{gugelmann2012random} this is called the global clustering coefficient. However, since this term is more often associated in the literature with the density of triangles compared to the number of paths of length 2, we use the term local clustering coefficient as is done, for instance, in~\cite{candellero2016clustering}.} is concentrated around its expectation which is shown to be $\bigT{1}$ as the number of vertices tends to infinity. In particular this results implies that the local clustering coefficient is asymptotically bounded away from zero. The convergence of this coefficient was not proven. In contrast, the global clustering coefficient was shown~\cite{candellero2016clustering} to converge in probability to some constant which can be explicitly stated as an integral expression, see Theorem 1.2 in \cite{candellero2016clustering}. The authors mention, however, that their analysis needs significant modification to be able to deal with convergence of the local clustering coefficient.

@@

\PvdH{I recall that J{\'u}lia told me that the results from~\cite{komjathy2018explosion} imply that the local clustering coefficient converges, although they do not mention it in the paper. Should we added a comment related to this?}

@@

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{figures/HyperRGG_N5000L5.eps}
\caption{Example of a Hyperbolic Random Graph $G_{\H,n}(\alpha, \nu)$ with $\alpha = ??$, $\nu = ??$ and $n = 5000$.
\PvdH{@Tobias: this figure (HyperRGG\_N5000L5.eps) is taken from the figures you send me. Could you fill in the value of $\alpha$ and $\nu$ you used for the simulation.}
}
\label{fig:H_graph_example}
\end{figure}

@@

Clustering measures for networks consider the fraction of triangles (triples of connected vertices) in the network. 
\MS{Do we need the definition of the global clustering coefficient? Also, I thought it was defined as 3 times number of triangles divided by number of length 2 paths and I do not see yet whether this agrees with the definition below.}
\PvdH{You were right. I have updated the definition.}
For instance, Let $G$ be a simple graph, with vertex set $V(G)$ and edge set $E(G)$, denote by $T_G$ and $\Lambda_G$, respectively, the total number of triangles and the total number of paths of length two in $G$. Then the \emph{global clustering coefficient} is defined as
\[
	c_G = \frac{3 \T_G}{\Lambda_G}.
\]
This coefficient measures the fraction of triangles in the network compared to the total possible number of triangles. A different global measure for clustering, \emph{local clustering coefficient} computes the average triangle density per node. To define this function, let $D_G(v)$ denote the degree of node $v$ and $N_G(k)$ the number of nodes with degree $k$ in the graph $G$. In addition let $\T_G(v,v_1,v_2) = \ind{(v,v_1) \in E} \ind{(v, v_2) \in E} \ind{(v_1, v_2) \in E}$ be the indicator that $v, v_1$ and $v_2$ form a triangle and write
\begin{equation}
	\T_G(v) = \sum_{v_1, v_2 \in V(G)} \T_G(v,v_1,v_2),
\end{equation}
to denote the number of triangles in which node $v$ participates. Then the \emph{local clustering coefficient} is given by
\[
	c_G = \frac{1}{|V(G)|} \sum_{v \in V(G)} \frac{T_G(v)}{\binom{D_G(v)}{2}}.
\]

However, since the local clustering coefficient assigns just one value to the whole network, representing its triangular structure, it is unable to characterize local structural properties involving triangles. The local clustering function, on the other hand, measures the fraction of triangles to which nodes of a given degree belong, compared to the maximum number of triangles in which they could participate~\cite{vazquez2002large,serrano2006clustering}. It therefore characterizes the triangular structure of nodes in the networks, based on their degree, and gives a more detailed look at the overall structure of the network. The formal definition of the \emph{local clustering function} is 
\begin{equation}\label{eq:def_local_clustering_general}
	c_{G}(k) = \begin{cases}
		\frac{1}{N_G(k)} \sum_{v \in V(G)}  \ind{D_G(v) = k} \frac{\T_G(v)}{\binom{k}{2}} &\mbox{if } N_G(k) \ge 1\\
		0 &\mbox{else.}
	\end{cases}
\end{equation}

%We note that other versions of the local clustering function have been considered, for instance where nodes with degree larger than $k$ are considered instead of nodes with degree exactly $k$. \PvdH{Motivate our choice for this local clustering function.}
